Ornaments as indicators of social changes and cultural practices in northeast Taiwan before and after European colonial period

### 2019-12-07 analysis/paper/paper.Rmd
fixed some typos and added sentences addressing the population issues

The island of Taiwan lies at the junction of mainland China, southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize the inherent biases in historical records and fragmented nature of texts selected by written record keepers that might lead to misinterpretations of artifacts [@Galloway2006]. Inconsistent and incomplete descriptions from different written sources lacking contexts are the problems need to be compared and examined by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeast Taiwan in early historical period, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metal working, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to early historical period was involved in complex exchange network both in a regional and global scale.

@@

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jar, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large amount of light gray glazed jars were found on the Spanish shipwreck San Diego sinking in 1600 AD, that is believed to be the earliest evidence yet found [@Hsieh2009; @Hsieh\\]. The origin of light gray glazed jars were thought from Southeast China and commonly found at sites in Taiwan that were associated with the European activities, such as Zeelandia fort site in Tainan [@Wang2007]. Regarding large dark brown glazed stoneware jars, they might be made in Southeast Asia and are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. Those jars might be directly imported to Kiwulan by the Europeans to Kiwulan or indirectly by Chinese merchants or other indigenous groups via the regional network in north Taiwan. Either way indicates that the Europeans played an important role for introducing those jars to Taiwan. I used the presence and absence of light grey glazed jar and large stoneware jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and the start of the European influence at Kiwulan.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened postholes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). These postholes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction across the sampling area. There are 30 burials in the sampling area account for one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes are 3,173. The use of ornaments as body adornments in burials will be discussed to understand the function of ornaments, but not statistically analysed since the amount of ornaments is highly affected by the presence of bead strands or patterned bands, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports the argument that ornaments were treated as high value items at Kiwulan based on a few finds in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place across different time periods as it can provide key information concerning social inequality indicated by uneven access to high-value item in relation to the foreign influences.

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells has less variety that the bells are typically >2 cm long with a wide variation of human faces as a motif. Although glass beads has less variety in size, small (0.5-1 cm) and medium (1-2 cm), they have wide variety of color or pattern mostly made by winding technique with high lead content in composition indicating possibly from China [@Cheng2008]. Despite more researches needed to understand the specific source of ornaments, current researches suggest that glass beads and metal ornaments have similar technique and composition to those found in China [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different periods in the upper component (1400-1900 AD). However, the glass beads found from lower component (700-1200 AD) are mostly Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC and declined at the early 2nd millennium [@Wang2018; @Francis2002]. It indicates a different bead source more related to the exchange network in the South China Sea for the lower component.

@@

Point pattern analysis can assess whether the distribution of artefacts represents hotspots produced by non-random processes [@Bevan2016; @Ducke2015], such as concentrations of ornaments resulting from social inequality stimulated by a colonial presence. To prepare the ornament location data for point pattern analysis, I assigned each ornament to a random coordinate pair located in 4 x 4 m square it was recovered from (ornaments lack individual point provenance). The next step was to divide the ornaments into three time periods. All squares in the sampling area were occupied during all phases of habitation based on the continuous presence of archaeological materials across over the strata representing a time span of 600 years. Although the quantity of locally made pottery declines in the later period with the growth of the Chinese porcelains, the general density of the artifacts present consistent thoughout all three phases.

Finally I computed the kernel densities for each time period for comparison. Kernel density estimations (KDE) compute the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here I use KDE to visualize core areas of ornaments and surrounding neighbourhoods [@Bonnier2019; @Cortegoso2016]. Density values of artefacts per square meter were calculated for each cell.

Figure \@ref(fig:plot-all-densities) shows that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period. There are three consistent sub-regions with a core area that shifts over time. The distribution might indicate increase and decrease in the number of different social groups who possessed more ornaments. The multiple groups during the European period might reflect more unequal consumption of ornaments across the site, relative to other periods, or random patterns resulting from bigger sample size. In addition, the generation of core areas might be biased due to small sample sizes, for example, a few ornaments found at one single square during the Chinese period could create an obvious hotspot.

@@

An indirect colonial influence is indicated at Kiwulan by a greater diversity of ornaments types and materials during the European period. Northeast Taiwan, Yilan, was involved in complex trading networks both in regional scale with other indigenous groups and Chinese merchants, and global scale with the European including the Dutch and the Spanish. The origin of ornaments might be multiple places mostly in Southeast Asia and introduced into northeast Taiwan by Chinese merchants before 17th century. Later, the trade activities became more frequent and intense in the 17th century that might be related to the colonial involvement of the Europeans based on the similar styles of golden beads found in Luzon that was once the colony of the Spanish during this period. In addition to ornaments, stonewares at Kiwulan that were also found on a Spanish shipwreck and Southeast Asia in relation to the Dutch and the Spanish trading routes, and the tobacco use were also believed introduced to Kiwulan by the Europeans [@Wang2007; @Chen2007]. The greater diversity and quantity of ornaments likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period in the early 19th century. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007]. Although historical records indicate that trade ornaments were prestige goods in prehistoric northeast Taiwan, the information might be biased that needs to be compared with archaeological evidence.

The archaeological contexts show that ornaments were rarely found in middens and mostly preserved well as body adornments in burial contexts serving as grave goods. This could reflect ornaments were viewed as valuable objects such as status indicators and their distribution across the dwelling area would help to understand the social relation reflected by people who possessed them, and the reaction of the society as a whole to the colonial presence in the 17th century. Spatial patterns of ornaments shows that their distribution was clustered during the pre-European and European period. Testing for spatial patterns indicates that these clusters are non-random, and are most highly concentrated during the European period. The spatial clustering during the pre-European and European period shows an increase in use of ornaments in the daily life at Kiwulan. It may indicate that a degree of social inequality based on uneven distribution was present before European contact and then it was reinforced and amplified during the European period. A further indicator of increasing social inequality is a burial dated to the 17th century that included 60 golden beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008].

@@

Another scenario could be that the increasing number and concentrated spatial pattern of ornaments indicate cultural resistance against the European intrusion by using them to show social identity and emphasize the local custom that had existed before European contact. Resistance could be presented in many forms. For example, @Rubertone discussed a mid-17th century Narragansett Indian cemetery where a large amount of native-made shell beads supposed to be tributes for the European were found in the graves as an act of political resistance against the colonial demands. Similarly, the frequent presence of ornaments across the dwelling area during the European suggests ornaments might be intentionally displayed in daily life to reinforce the original cultural tradition. During the Chinese period in the 19th century, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This might reflect the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals to accumulate wealth, or population decrease. It is also possible that some ornaments were treated as heirlooms, such as carnelian beads and golden beads, that passed from one generation to the next. This might suggest that the ornaments found in Chinese period were old item continuously used until then. Further analysis about the source for ornaments from different periods will be helpful to understand their origin and reginal circulation.

>>>>

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered the historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in the Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize the inherent biases in historical records and the fragmented nature of texts selected by written record keepers that might lead to misinterpretations of artifacts [@Galloway2006]. Inconsistent and incomplete descriptions from different written sources lacking contexts are the problems need to be compared and examined by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeast Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, a large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to the early historical period was involved in complex exchange networks both on a regional and global scale.

@@

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. A large amount of light gray glazed jars were found on the Spanish shipwreck San Diego sinking in 1600 AD, which is believed to be the earliest evidence yet found [@Hsieh2009; @Hsieh\\]. The origin of light gray glazed jars was thought from Southeast China and commonly found at sites in Taiwan that were associated with the European activities, such as Zeelandia fort site in Tainan [@Wang2007]. Regarding large dark brown glazed stoneware jars, they might be made in Southeast Asia and are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. Those jars might be directly imported to Kiwulan by the Europeans to Kiwulan or indirectly by Chinese merchants or other indigenous groups via the regional network in north Taiwan. Either way indicates that the Europeans played an important role in introducing those jars to Taiwan. I used the presence and absence of light grey glazed jar and large stoneware jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and the start of the European influence at Kiwulan.

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened postholes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). These postholes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction across the sampling area. There are 30 burials in the sampling area account for one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes is 3,173. The use of ornaments as body adornments in burials will be discussed to understand the function of ornaments, but not statistically analysed since the number of ornaments is highly affected by the presence of bead strands or patterned bands, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports the argument that ornaments were treated as high-value items at Kiwulan based on a few finds in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place across different time periods as it can provide key information concerning social inequality indicated by uneven access to high-value items in relation to the foreign influences.

@@

The distribution of frequencies for subtypes in each major class is presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, small (0.5-1 cm) and medium (1-2 cm), they have a wide variety of colors or patterns mostly made by winding technique with high lead content in composition indicating possibly from China [@Cheng2008]. Despite more researches needed to understand the specific source of ornaments, current researches suggest that glass beads and metal ornaments have similar technique and composition to those found in China [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different periods in the upper component (1400-1900 AD). However, the glass beads found from the lower component (700-1200 AD) are mostly Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC and declined in the early 2nd millennium [@Wang2018; @Francis2002]. It indicates a different bead source more related to the exchange network in the South China Sea for the lower component.

@@

Point pattern analysis can assess whether the distribution of artefacts represents hotspots produced by non-random processes [@Bevan2016; @Ducke2015], such as concentrations of ornaments resulting from social inequality stimulated by a colonial presence. To prepare the ornament location data for point pattern analysis, I assigned each ornament to a random coordinate pair located in 4 x 4 m square it was recovered from (ornaments lack individual point provenance). The next step was to divide the ornaments into three time periods. All squares in the sampling area were occupied during all phases of habitation based on the continuous presence of archaeological materials across over the strata representing a time span of 600 years. Finally I computed the kernel densities for each time period for comparison. Kernel density estimations (KDE) compute the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here I use KDE to visualize core areas of ornaments and surrounding neighbourhoods [@Bonnier2019; @Cortegoso2016]. Density values of artefacts per square meter were calculated for each cell.

Figure \@ref(fig:plot-all-densities) shows that there is one major core area during the pre-European period, multiple core areas during the European period, and a single core during the Chinese period. There are three consistent sub-regions with a core area that shifts over time. The distribution might indicate increase and decrease in the number of different social groups who possessed more ornaments. The multiple groups during the European period might reflect more unequal consumption of ornaments across the site, relative to other periods, or random patterns resulting from a bigger sample size. In addition, the generation of core areas might be biased due to small sample sizes, for example, a few ornaments found at one single square during the Chinese period could create an obvious hotspot. The proportion of ornaments is relative to the number of locally made pottery that shows the pattern of decline in the 19th century. Although the quantity of locally made pottery declines in the 19th century with an obvious growth of the imported Chinese porcelains, the general density of the total number of artifacts is consistent throughout all three phases. However, less locally made pottery and ornaments might relate to the indigenous population decline. Whether the observed clustering is random or non-random is crucial for the interpretation of intentional human activities.

@@

An indirect colonial influence is indicated at Kiwulan by a greater diversity of ornament types and materials during the European period. Northeast Taiwan, Yilan, was involved in complex trading networks both in regional scale with other indigenous groups and Chinese merchants, and global scale with the European including the Dutch and the Spanish. The origin of ornaments might be multiple places mostly in Southeast Asia and introduced into northeast Taiwan by Chinese merchants before 17th century. Later, the trade activities became more frequent and intense in the 17th century that might be related to the colonial involvement of the Europeans based on the similar styles of golden beads found in Luzon that was once the colony of the Spanish during this period. In addition to ornaments, stonewares at Kiwulan that were also found on a Spanish shipwreck and Southeast Asia in relation to the Dutch and the Spanish trading routes, and the tobacco use were also believed introduced to Kiwulan by the Europeans [@Wang2007; @Chen2007]. The greater diversity and quantity of ornaments likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period in the early 19th century. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007]. Although historical records indicate that trade ornaments were prestige goods in prehistoric northeast Taiwan, the information might be biased that needs to be compared with archaeological evidence.

The archaeological contexts show that ornaments were rarely found in middens and mostly preserved well as body adornments in burial contexts serving as grave goods. This could reflect ornaments were viewed as valuable objects such as status indicators and their distribution across the dwelling area would help to understand the social relation reflected by people who possessed them, and the reaction of the society as a whole to the colonial presence in the 17th century. Spatial patterns of ornaments show that their distribution was clustered during the pre-European and European period. Testing for spatial patterns indicates that these clusters are non-random, and are most highly concentrated during the European period. The spatial clustering during the pre-European and the European period shows an increase in the use of ornaments in daily life at Kiwulan. It may indicate that a degree of social inequality based on uneven distribution was present before European contact and then it was reinforced and amplified during the European period. A further indicator of increasing social inequality is a burial dated to the 17th century that included 60 golden beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008].

@@

Another scenario could be that the increasing number and concentrated spatial pattern of ornaments indicate cultural resistance against the European intrusion by using them to show social identity and emphasize the local custom that had existed before European contact. Resistance could be presented in many forms. For example, @Rubertone2000 discussed a mid-17th century Narragansett Indian cemetery where a large amount of native-made shell beads supposed to be tributes for the European were found in the graves as an act of political resistance against the colonial demands. Similarly, the frequent presence of ornaments across the dwelling area during the European suggests ornaments might be intentionally displayed in daily life to reinforce the original cultural tradition. During the Chinese period in the 19th century, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This might reflect the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals to accumulate wealth, or population decrease. It is also possible that some ornaments were treated as heirlooms, such as carnelian beads and golden beads, that passed from one generation to the next. This might suggest that the ornaments found in the Chinese period were old items continuously used until then. Further analysis of the source for ornaments from different periods will be helpful to understand their origin and regional circulation.
